Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations

### 2022-11-08 paper/paper/SPIE/article.tex
[WIP]

\documentclass[12pt]{spieman}  % 12pt font required by SPIE;
%\documentclass[a4paper,12pt]{spieman}  % use this instead for A4 paper
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{graphicx}
\graphicspath{ {./figures/} }
\usepackage{setspace}
\usepackage{tocloft}
\usepackage{doi}

\newcommand{\code}[1]{\small \texttt{#1} \normalsize}

@@

\section{Methods}
\label{sect:methods}

\subsection{fMRI Preprocessing}

@@

The time and space complexities of computing the full \(N \times N\) correlation
matrix of \(\mathbf{M}\), and the subsequent eigenvalues, are too large to be
tractable. However, since transposition does not change the eigenvalues, we
can use the transpose to efficiently compute these eigenvalues (see
Appendix \ref{sec:transpose}).

@@

\subsubsection{Non-RMT Eigenfeatures}
\label{sec:eigs-only}

@@

@@

It could easily be the case that a combination of RMT and non-RMT eigenfeatures
are more predictive than either alone. We thus test a variety of eigenfeature
combinations, with the motivation for these particular combinations being to
test whether RMT eigenfeatures add anything beyond smoothing.
...

@@

MAYBE MOVE THIS SECTION AFTER RMT AND EIGEN FEATURES?

@@

Baseline reductions use as statistical summaries: robust measures of location
(mean, max, min); non-robust measures of location (median, \(95^{\text{th}}\)
percentile, \(5^{\text{th}}\) percentile); non-robust measures of scale
(standard deviation, range); and robust measures of scale (interquartile range,
difference between \(95^{\text{th}}\) percentile and \(5^{\text{th}}\)
percentile). These statistics are used to summarizing / reduce the fMRI data
along the voxel dimension, yielding a feature of \(T\) dimensions. E.g. the "mean"
feature is just the global mean signal. Exact details can be found in the
\href{https://github.com/DM-Berger/random-matrix-fmri/blob/master/code/rmt/enumerables.py#L51-L69}{paper
source code}. Note we do not take slice variants of these baseline features.

\subsubsection{Final Feature List}

[INSERT TABLE OF ALL FEATURES]

@@

Our investigation can be structured hierarchically: we have a number of
\textit{datasets}, and each dataset has one or more \textit{classification
tasks}. Each classification task can be solved with a \textit{classifier}.
Each classifier fit operates on a \textit{featyre}, and each feature is
extracted via a certain \textit{preprocessing degree} and various
\textit{analytic choices} (slicing, trimming, unfolding degree, normalization).

>>>>

\documentclass[12pt]{spieman}  % 12pt font required by SPIE;
%\documentclass[a4paper,12pt]{spieman}  % use this instead for A4 paper
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{graphicx}
\graphicspath{ {./figures/} }
\usepackage{setspace}
\usepackage{tocloft}
\usepackage{doi}
\usepackage{csquotes}
\MakeOuterQuote{"}

\newcommand{\code}[1]{\small \texttt{#1} \normalsize}
\newcommand{\tcode}[1]{\footnotesize \texttt{#1} \normalsize}

@@

\section{Methods}
\label{sec:methods}

\subsection{fMRI Preprocessing}
\label{sec:preproc}

@@

The time and space complexities of computing the full \(N \times N\)
correlation matrix of \(\mathbf{M}\), and the subsequent eigenvalues, are too
large to be tractable. However, since transposition does not change the
eigenvalues, we can use the transpose to efficiently compute these eigenvalues
(see Appendix \ref{sec:transpose}). These raw eigenvalues (feature \code{eigs}
in Table \ref{tab:features}), most directly test the idea that eigenvalues
alone are useful functional connectivity features.

However, not all eigenvalues may have predictive utility. We also examine as
features the central eigenvalues (defined as the middle 10\%, 20\%, or 40\% of
the sorted eigenvalues, \code{eigsmiddle} in Table \ref{tab:features}), and
tail eigenvalues (defined as the first and last 5\%,  10\%, and 20\% of each
tail of the spectrum, \code{eigsminmax} in Table \ref{tab:features}).

@@

\subsubsection{Smoothed Eigenfeatures}
\label{sec:eigs-only}

@@

It could easily be the case that a combination of RMT and non-RMT eigenfeatures
are more predictive than either alone. We thus test a variety of eigenfeature
combinations, with the motivation for these particular combinations being to
test whether RMT eigenfeatures add anything beyond smoothing. Combined features
are formed by concatenation, so that if we have features \(f_1, \dots,  f_n\)
with dimensions \(p_1, \dots, p_n\), then the combined feature is \([f_1;
\dots; f_n]\) with \(\sum p_i\) dimensions.  The final combinations chosen can
be found in Table \ref{tab:features}.

@@

For combined features, slicing variants are also computed, with slicing
performed first on each feature to be combined. E.g. if we have features \(f_1,
\dots,  f_n\) with dimensions \(p_1, \dots, p_n\), then the sliced max 25\%
feature for each component feature, \(\hat{f}_i\) is the last \(p_i / 4\)
elements of \(f_i\), and the combined sliced feature is  \([\hat{f}_1;\;
\dots;\; \hat{f}_n]\)\footnote{Exact implementation can be found in the
\href{https://github.com/DM-Berger/random-matrix-fmri/blob/7c9e4187f582dedee728cd7193b8894d928c2f00/code/rmt/updated\_predict.py\#L182-L213}{source code}}.

@@

@@

Baseline reductions use as statistical summaries: robust measures of location
(mean, max, min); non-robust measures of location (median, \(95^{\text{th}}\)
percentile, \(5^{\text{th}}\) percentile); non-robust measures of scale
(standard deviation, range); and robust measures of scale (interquartile range,
difference between \(95^{\text{th}}\) percentile and \(5^{\text{th}}\)
percentile). These statistics are used to summarizing / reduce the fMRI data
along the voxel dimension, yielding a feature of \(T\) dimensions. E.g. the ``mean''
feature is just the global mean signal. Exact details can be found in the
\href{https://github.com/DM-Berger/random-matrix-fmri/blob/master/code/rmt/enumerables.py#L51-L69}{paper
source code}.

We do not take slice variants of these baseline features, since the baseline
features are still time series. However, we do also evaluate a number of
smoothing degrees of these features, to account for noise and to be somewhat
similar to the various smoothing variants for the eigenfeatures. Each baseline
feature thus has a uniformly smoothed variant, where the size of the smoothing
window is either 1 (no smoothing), 2, 4, 8, or 16.

\subsection{Classifiers}

We use a variety of stadard machine-learning classifiers available in the
Scikit-learn\cite{pedregosaScikitlearnMachineLearning2011} Python library to
solve each classification task. We use a gradient-boosted decision tree (GBDT),
random forest (RF) classifier, support vector classifier (SVC) with radial
basis function, and k-nearest neighbours classifiers with k equal to 3, 5, and
9 (KNN3, KNN5, and KNN9, respectively). We originally also attempted to test a
simple logistic regression classifier, but found that this model to frequently
failed to converge for a number of features, even after increasing iterations
significantly. In all cases the default hyperparameter values were used.

\subsection{Preprocessing Levels}

Each preprocessing intermediate in the (1) brain extraction, (2) slice time correction,
(3) motion correction, and (4) template registration pipeline was saved and used for
entirely separate feature analyses. E.g. all features were extracted from a brain
prepared with one of four levels or degrees of preprocessing, where level \(k\)
includes preprocessings steps 1 to \(k\), inclusive, and starting at \(k = 1\).

\begin{table}[h!]
\caption{
    Feature groupings for summarization. eigs = non-RMT eigenfeatures and their combinations with
    other non-RMT eigenfeatures. rmt = RMT eigenfeatures and their combinations. tseries = baseline
    timeseries reductions.\\
}
\label{tab:features}
\small
\centering
\begin{tabular}{lllr}
\hline
\textbf{Gross Source} & \textbf{Feature Group} & \textbf{Feature id}   \\
\hline
eigs           & eigs          & \tcode{eigs}                           \\
               & eigs max      & \tcode{eigsminmax10}                   \\
               &               & \tcode{eigsminmax20}                   \\
               &               & \tcode{eigsminmax5}                    \\
               & eigs middle   & \tcode{eigsmiddle10}                   \\
               &               & \tcode{eigsmiddle20}                   \\
               &               & \tcode{eigsmiddle40}                   \\
               & eigs smooth   & \tcode{eigs + eigs\_smooth}            \\
               &               & \tcode{eigs + savgol}                  \\
               &               & \tcode{eigs\_savgol}                   \\
               &               & \tcode{eigs\_smooth}                   \\
               &               &                                        \\
rmt            & rmt + eigs    & \tcode{eigs + levelvar}                \\
               &               & \tcode{eigs + rigidity}                \\
               &               & \tcode{eigs + rigidity + levelvar}     \\
               &               & \tcode{eigs + unfolded}                \\
               &               & \tcode{eigs + unfolded + levelvar}     \\
               &               & \tcode{eigs + unfolded + rigidity}     \\
               & rmt only      & \tcode{levelvar}                       \\
               &               & \tcode{rigidity}                       \\
               &               & \tcode{rigidity + levelvar}            \\
               &               & \tcode{unfolded}                       \\
               &               & \tcode{unfolded + levelvar}            \\
               &               & \tcode{unfolded + rigidity}            \\
               &               & \tcode{unfolded + rigidity + levelvar} \\
               &               &                                        \\
tseries        & location      & \tcode{T-max}                          \\
               &               & \tcode{T-mean}                         \\
               &               & \tcode{T-med}                          \\
               &               & \tcode{T-min}                          \\
               &               & \tcode{T-p05}                          \\
               &               & \tcode{T-p95}                          \\
               & scale         & \tcode{T-iqr}                          \\
               &               & \tcode{T-rng}                          \\
               &               & \tcode{T-rrng}                         \\
               &               & \tcode{T-std}                          \\
\hline
\end{tabular}
\end{table}

@@

We perform a \textit{multiverse
analysis}\cite{steegenIncreasingTransparencyMultiverse2016} to assess the
overall predictive potential of the various eigenfeatures across all
previously-mentioned analytic choices. That is, for each combination of
comparison task and classifier, we use 5-fold cross validation, and use the
mean AUROC (mAUROC) across folds as our metric to evaluate feature predictive utility.

Each non-baseline feature is evaluated, for each classification task and
classifier, exactly 1280 times: there are 4 preprocessing levels \(\times\)
2 normalization methods \(\times\) 4 trimming choices \(\times\) 4 unfolding
degrees  \(\times\) 10 slice choices. Each baseline feature is evaluated
exactly 40 times for each comparison task and classifier: 4 preprocessing
levels \(\times\) 2 normalization methods \(\times\) 5 smoothing window sizes.
In total, with all the datasets examined in this study, this yields 2 053 920
mAUROC values to summarize, with 5 main \textit{analytic factors}:
preprocessing, normalization, trimming, degree, and slicing.

\section{Results}

Summarizing this amount of data requires some caution. Measures of location or
scale, even when robust, are, for the most part, misleading and uninformative when distributions
are skewed, and summarizing too broadly is likewise uninformative. E.g. the
average mAUROC across all over \(2 \times 10^6\) model evaluations is too
broad a summary to meaningfully interpret. Likewise, each analytic factor
could produce an interaction: e.g. perhaps
a meaningful number.

Our multiverse analysis can be structured semi-hierarchically, where choices at each level
in the hierarchy impact the final distribution of AUROCs observed. We have a
number of \textit{datasets}, and each dataset has one or more
\textit{classification tasks}. Each classification task can be solved with a
\textit{classifier}. Each classifier fit operates on a \textit{feature}, and
each feature is extracted via a certain \textit{preprocessing degree} and
various \textit{analytic choices} (slicing, trimming, unfolding degree,
normalization).
